$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-04 paper/paper.tex
Comment on the work by Simon Gay

Considerable efforts in the mechanisation of resource-aware calculi have been made in recent years.

Leftover typing as presented in this paper is based on work by Allais \cite{Allais2018a}.
Allais uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
This type system sits on top of an untyped but well-scoped syntax with de Bruijn indices.
He proves type preservation and provides a decision procedure for type checking and type inference.
In this paper, we apply his technique to the \picalc{}, but limit ourselves to type-checking (scope restriction needs usage annotations) terms, and focus on generalising the usage algebra.

@@

The work by Goto et al. \cite{Goto2016a} is likely the first contribution to session-types that comes along with a mechanised proof of type safety (in Coq). 
Their paper extends session types with polymorphism and pattern matching on session types.
They use a locally-nameless encoding of variable references, a syntax prior to types, and a LTS semantics that encodes session-typed processes into the \picalc{}.
Their type system uses reorering of contexts and extrinsic contex splits.

\cite{Rouvoet2020}
\cite{Veltri2020}
\cite{Gay2001}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future Work}

Work that will be done time permitting:

Erika Kreuer: model usage algebra as a state machine?

Adding bidirectionality would provide us with a decision procedure for type checking.

\begin{description}

\item [Proof of progress]

\item [Product types]

\item [Sum types]

\item [Decidable typechecking]

  Bidirectional typing rules needed.

\item [Soundness and completeness with respect to an alternative formalization.]

\item [Encoding of session types]

\end{description}

>>>>

Leftover typing as presented in this paper is based on work by Allais \cite{Allais2018a}.
Allais uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
This type system sits on top of an untyped but well-scoped syntax with de Bruijn indices.
He proves type preservation and provides a decision procedure for type checking and type inference.
In this paper, we apply his technique to the \picalc{}, but limit ourselves to type-checking (scope restriction needs usage annotations) terms, and focus on generalising the usage algebra.
Drawing from quantitative type theory by McBride and Atkey \cite{McBride2016, Atkey2018}, it is crucial for us too to be able to talk about fully consumed resources.
As a result, we are able to transmit fully exhausted channels over channels.

Recent years have seen an increase in the efforts to mechanise resource-aware concurrency calculi, but the formalisation of the linear \picalc{} in Isabelle/HOL by Gay \cite{Gay2001} is perhaps one of the earliest works.
Gay encodes the \picalc{} with linear and shared types into a more general framework.
As we do, he uses de Bruijn indices, a reduction relation and a type system posterior to the syntax.
However, in his work context splits are explicit, and the types of used multiplicities erased.

@@

The work by Goto et al. \cite{Goto2016a} is likely the first formal contribution to session-types that comes along with a mechanised proof of type safety (in Coq). 
Their paper extends session types with polymorphism and pattern matching on session types.
They use a locally-nameless encoding of variable references, a syntax prior to types, and a LTS semantics that encodes session-typed processes into the \picalc{}.
Their type system uses reorering of contexts and extrinsic contex splits.

\cite{Rouvoet2020}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

Mechanisation of the linear, gradual and shared \picalc{} in less than 1500 lines of code.

Adding product types and sum types.
Proving that progress preserves the balancing of channels.
Erika Kreuer has suggested modeling the usage algebra as a state machine.
Adding bidirectionality would provide us with a decision procedure for type checking for a given context of algebras.
